
```{r}
library(readr)
library(glmnet)
library(grpreg)
library(rsample)
library(ggplot2)
library(dplyr)
library(tidyr)
library(tibble)
```

```{r}
fuels_data <- read_csv("fuels_data.csv", col_types = cols(site = col_factor()))
head(fuels_data)
```

```{r}
## training and testing data
set.seed(230)
split_data <- initial_split(fuels_data, prop = 0.75, strata = site)
train_data <- training(split_data)
test_data <- testing(split_data)
```

```{r}
fit_model <- function(train_df, test_df, model_type) {
  
  # create model matrices 
  x_train <- model.matrix(fuel_load_Mg_ha ~ ., data = train_df)[,-1]
  x_test <- model.matrix(fuel_load_Mg_ha ~ ., data = test_df)[,-1]
  
  y_train <- train_df$fuel_load_Mg_ha
  y_test <- test_df$fuel_load_Mg_ha
  
  if(model_type == "ridge") {
    
    # ridge regression with cross-validation 
    cv_fit <- cv.glmnet(x = x_train, y = y_train, alpha = 0, type.measure = "mse", standardize = TRUE)
    
    # test and train set predictions 
    y_hat_train <- as.numeric(predict(cv_fit, newx = x_train, s = "lambda.min"))
    y_hat_test <- as.numeric(predict(cv_fit, newx = x_test, s = "lambda.min"))
    
  } else if (model_type == "lasso") {

    # lasso regression with cross-validation 
    group <- c(rep(1, 5), seq_len(ncol(x_train) - 5) + 1)
    cv_fit <- cv.grpreg(X = x_train, y = y_train, group = group, penalty = "grLasso")
    coefs <- coef(cv_fit, lambda = cv_fit$lambda.min)
    nonzero_coef_names <- setdiff(names(coefs)[coefs != 0], "(Intercept)")

    # test and train set predictions 
    y_hat_train <- as.numeric(predict(cv_fit, X = x_train, lambda = cv_fit$lambda.min))
    y_hat_test <- as.numeric(predict(cv_fit, X = x_test, lambda = cv_fit$lambda.min))
    
    
    # coefficient path plot 
    fit <- cv_fit$fit
    beta_matrix <- as.matrix(coef(fit))
    
    coef_path_df <- as.data.frame(beta_matrix) %>%
      rownames_to_column("term")

    colnames(coef_path_df)[-1] <- fit$lambda
    
    coef_path_df_long <- coef_path_df %>%
      pivot_longer(cols = -term,
                   names_to = "lambda",
                   values_to = "estimate") %>%
      mutate(lambda = as.numeric(lambda)) %>%
      filter(term != "(Intercept)",
             lambda <= 10)
    
    coef_plot <- ggplot(coef_path_df_long, aes(x = lambda, y = estimate, group = term, color = term)) +
      geom_line(linewidth = 0.5) +
      geom_vline(xintercept = cv_fit$lambda.min, linetype = "dashed", linewidth = 0.8) +
      theme_minimal() +
      theme(legend.position = "none",
            panel.grid.minor = element_blank(),
            panel.border = element_rect(color = "black", fill = NA, linewidth = 1.2),
            axis.title = element_text(size = 14),
            axis.text = element_text(size = 12)) +
      labs(x = expression(lambda),
           y = "Coefficient") + 
      scale_x_continuous(limits = c(0, 10), breaks = 0:10)
    
  }
  
  # test-set performance
  rmse <- sqrt(mean((y_test - y_hat_test)^2))
  
  rss <- sum((y_test - y_hat_test)^2)
  tss <- sum((y_test - mean(y_test))^2)
  r2 <- (1-(rss/tss))

  # diagnostic plots
  plot_df <- data.frame(
    y = y_train,
    y_hat = y_hat_train,
    resid = y_train - y_hat_train
  )

  resid_plot <- ggplot(plot_df, aes(x = y_hat, y = resid)) +
    geom_hline(yintercept = 0, color = "grey40", linetype = "dashed") +
    geom_point(alpha = 0.4) +
    geom_smooth(method = "loess", span = 1, se = FALSE, color = "brown3") + 
    labs(x = "Predicted value",
         y = "Residual") +
    theme_minimal() + 
    theme(panel.grid.minor = element_blank(),
          panel.border = element_rect(color = "black", fill = NA, linewidth = 1.2),
          axis.title = element_text(size = 14),
          axis.text = element_text(size = 12))
  
  fitted_plot <- ggplot(plot_df, aes(x = y, y = y_hat)) +
    geom_point(alpha = 0.4) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey40") + 
    labs(x = "Measured surface fuel load (Mg/ha)",
         y = "Predicted surface fuel load (Mg/ha)") +
    theme_minimal() + 
    theme(panel.grid.minor = element_blank(),
          panel.border = element_rect(color = "black", fill = NA, linewidth = 1.2),
          axis.title = element_text(size = 14),
          axis.text = element_text(size = 12)) + 
    scale_y_continuous(limits = c(0, 300)) + 
    scale_x_continuous(limits = c(0, 300))
  
  if(model_type == "ridge") {
    
    return(list(RMSE = rmse, R2 = r2, residual_plot = resid_plot, fitted_line_plot = fitted_plot))
    
  } else if (model_type == "lasso") {
    
    return(list(RMSE = rmse, R2 = r2, residual_plot = resid_plot, fitted_line_plot = fitted_plot, 
                coef_path_plot = coef_plot, selected_coefs = nonzero_coef_names))
    
  }
  
}
```

```{r}
ridge <- fit_model(train_df = train_data, test_df = test_data, model_type = "ridge")
ridge$RMSE
ridge$R2

#png(file = "report_figs/ridge_resid.png", width = 7.5*600, height = 5*600, res = 600)
ridge$residual_plot
#dev.off()

#png(file = "report_figs/ridge_fitted_line.png", width = 6*600, height = 4*600, res = 600)
ridge$fitted_line_plot
#dev.off()
```

```{r}
lasso <- fit_model(train_df = train_data, test_df = test_data, model_type = "lasso")
lasso$RMSE
lasso$R2
lasso$selected_coefs

#png(file = "report_figs/lasso_resid.png", width = 7.5*600, height = 5*600, res = 600)
lasso$residual_plot
#dev.off()

#png(file = "report_figs/lasso_coefs.png", width = 7.5*600, height = 5*600, res = 600)
lasso$coef_path_plot
#dev.off()
```

# Conformal Prediction

```{r}

fit_model_with_conformal <- function(train_df, test_df, model_type) {
  
  # create model matrices 
  x_train <- model.matrix(fuel_load_Mg_ha ~ ., data = train_df)[,-1]
  x_test <- model.matrix(fuel_load_Mg_ha ~ ., data = test_df)[,-1]
  
  y_train <- train_df$fuel_load_Mg_ha
  y_test <- test_df$fuel_load_Mg_ha
  
  if(model_type == "ridge") {
    
    # ridge regression with cross-validation 
    cv_fit <- cv.glmnet(x = x_train, y = y_train, alpha = 0, type.measure = "mse", standardize = TRUE)
    
    # test and train set predictions 
    y_hat_train <- as.numeric(predict(cv_fit, newx = x_train, s = "lambda.min"))
    y_hat_test <- as.numeric(predict(cv_fit, newx = x_test, s = "lambda.min"))
  
    # test-set performance
    rmse <- sqrt(mean((y_test - y_hat_test)^2))
    
    rss <- sum((y_test - y_hat_test)^2)
    tss <- sum((y_test - mean(y_test))^2)
    r2 <- (1-(rss/tss))
    
    # conformal prediction add-on
    lambda_min <- cv_fit$lambda.min #fix lambda and accept approx
    n_tr  <- nrow(x_train)
    p_tr  <- ncol(x_train)
    X_aug <- cbind(1, x_train) # add intercept
    
    beta_vec <- as.numeric(coef(cv_fit, s = "lambda.min")) # return sparse matrix
    Resvec   <- as.numeric(y_train - X_aug %*% beta_vec) # residuals manual
    e_sigma  <- sqrt(sum(Resvec^2) / (n_tr - p_tr - 1)) # residual sd
    
    # hat matrix needs intercept-augmented gram
    Gram_inv_aug <- solve(t(X_aug) %*% X_aug + lambda_min * diag(p_tr + 1)) # penalized version compared to OLS
    H        <- X_aug %*% Gram_inv_aug %*% t(X_aug)
    leverage <- diag(H)
    resmat   <- diag(n_tr) - H
    
    grid.y <- seq(min(y_train) - 200, max(y_train) + 200, by = 0.1) # we can adjsut these
    cvr    <- ceiling(0.95 * (n_tr + 1))   # conformal rank threshold
    
    loo_pred <- matrix(0, n_tr, 5)
    loo_cov  <- matrix(0, n_tr, 2)
    
    for (i in seq_len(n_tr)) {
      
      # LOO coefficients via Sherman-Morrison (approximate for ridge)
      beta_i  <- beta_vec - Gram_inv_aug %*% X_aug[i, ] * Resvec[i] / (1 - leverage[i])
      #e_sigma_i <- sqrt(e_sigma^2 * (n_tr - p_tr) - Resvec[i]^2 / (1 - leverage[i])) / sqrt(n_tr - p_tr - 1)
      pred_i  <- sum(X_aug[i, ] * beta_i)
      
      # conformal interval
      grid_r <- sapply(grid.y, function(y) {
        Res <- Resvec + resmat[, i] * (y - y_train[i])
        rank(abs(Res))[i]
      })
      Cinterval        <- range(grid.y[grid_r <= cvr])
      loo_pred[i, 4:5] <- Cinterval
      loo_cov[i, 2]    <- findInterval(y_train[i], Cinterval)
    }
    
    conformal_intervals_df = data.frame(
        y_true    = y_train,
        lower     = loo_pred[, 4],
        upper     = loo_pred[, 5],
        width     = loo_pred[, 5] - loo_pred[, 4],
        covered   = loo_cov[, 2] == 1
      )
    
    # plotting stuff
    plot_df <- data.frame(
      y     = y_train,
      y_hat = y_hat_train,
      resid = y_train - y_hat_train,
      lower = conformal_intervals_df$lower,
      upper = conformal_intervals_df$upper,
      covered = conformal_intervals_df$covered
  )
    
    fitted_plot_intervals <- ggplot(plot_df, aes(x = y, y = y_hat)) +
      geom_ribbon(aes(ymin = lower, ymax = upper), 
                  fill = 'steelblue', alpha = 0.2) +
      geom_point(alpha = 0.4) +
      geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey40") + 
      labs(x = "Measured surface fuel load (Mg/ha)",
           y = "Predicted surface fuel load (Mg/ha)") +
      theme_minimal() + 
      theme(panel.grid.minor = element_blank(),
            panel.border = element_rect(color = "black", fill = NA, linewidth = 1.2),
            axis.title = element_text(size = 14),
            axis.text = element_text(size = 12)) + 
      scale_y_continuous(limits = c(0, 500)) + 
      scale_x_continuous(limits = c(0, 300))
    
    return(list(
      RMSE = rmse,
      R2 = r2,
      fitted_line_plot_intervals = fitted_plot_intervals,
      conformal_coverage = mean(loo_cov[, 2] == 1),
      # needed stuff for plotting?
      conformal_intervals = conformal_intervals_df)
    )
  }
}
```

```{r}
ridge <- fit_model_with_conformal(train_df = train_data, test_df = test_data, model_type = "ridge")

```

```{r}
ridge$conformal_coverage
ridge$conformal_intervals

```

```{r}
ridge$fitted_line_plot_intervals
ridge$R2
```

in OLS there is no penalty so dropping an observation doesn't change the optimization problem structure, just the data. With ridge, the proper procedure would re run cross validation each time to find a new optimal lambda wihtout observation i but that would be a lot of runs so we just fix lambda at the full data withthe minimm

we need to add X_aug with an intercept column because we are doing matrix algebra by hand  and we want it to be part of the computation, but we dont want it to be shrunk towards zero- glm net handles it internally so we add it back here
