
## Data and Packages
```{r warning = FALSE, message = FALSE}
# packages
library(tidyverse)
library(tidymodels)
library(probably)
```

```{r warning = FALSE, message = FALSE}
# data
fuels_data <- readr::read_csv("fuels_data.csv")
fuels_data$site <- factor(fuels_data$site, levels = c("BLG","DLB", "IND","SHA","TCU","WIN"))
  
set.seed(24)
fd_split <- initial_split(fuels_data, prop=0.75, strata = site )
fd_train <- training(fd_split)
fd_test <- testing(fd_split)
```

## Set up Ridge Model

```{r}
# create model
ridge_model <- linear_reg(mode = "regression", engine = "glmnet",
                          penalty = tune(),
                          mixture = 0) # make this 1 if we want lasso
ridge_wflow <- workflow() |>
  add_model(ridge_model)
```

```{r}
# add recipe
ridge_recipe <- recipe( fuel_load_Mg_ha ~., data = fd_train) |>
  step_rm(MinSD) |>
  step_zv(all_numeric_predictors()) |>
  step_nzv(all_numeric_predictors()) |> 
  step_novel(all_nominal_predictors()) |>
  step_normalize(all_numeric_predictors()) |>
  step_dummy(all_nominal_predictors())

ridge_wflow <- ridge_wflow |>
  add_recipe(ridge_recipe)
```

```{r}
# tune the model 
set.seed(432)
fd_cv <- vfold_cv(fd_train, v = 10) # we can adjust this to be 5 fold too

ridge_tune1 <- tune_grid(ridge_model,
                         ridge_recipe,
                         resamples = fd_cv,
                         grid = grid_regular(penalty(range = c(-5, 5)), levels = 50))
```

## exploration of warnings
```{r}
# zero variance in MinSD
unique(fd_train$MinSD)
```
all values of MinSD are zero, causing issues so we could probably remove this too; adding step_rm(MinSD) as a fix

```{r}
# unseen Plot levels
fd_train |> count(plot) |> arrange(n)
```
most plots only have one observation in them, so creating very sparse dummy matrix- may be helpful to remove this variable all together- adding step_novel to gracefully handle new levels

```{r}
# correlation computation, estimate constand and 0 SD
# check variance of all numeric predictors
fd_train |>
  select(where(is.numeric)) |>
  summarise(across(everything(), var, na.rm = TRUE)) |>
  pivot_longer(everything(), names_to = "variable", values_to = "variance") |>
  arrange(variance)
```
will probably be easiest to handle this with step_zn (removes any perfectly constant predictor (minSD)) and stem_nzv (almost constant) instead of going through each variable individually

## back to modeling

```{r}
# plotting penalty values
ridge_tune1 |>
  collect_metrics() |>
  filter(.metric == "rmse") |>
  ggplot(mapping = aes(x = penalty, y = mean)) + geom_point() + geom_line() +
  scale_x_log10() #because penalty is on the log scale
```

we want to pick the penalty value that minimizes RMSE (we can change this if we want to some other metric) 

```{r}
# select best ridge
ridge_best <- ridge_tune1 |>
  select_best(metric = "rmse")

ridge_best

ridge_wflow_final <- ridge_wflow |>
  finalize_workflow(parameters = ridge_best)
```

```{r}
# ridge plot check with calibration
ridge_pred_check <- ridge_wflow_final |>
  fit_resamples(
    resamples = fd_cv,
    control = control_resamples(save_pred = TRUE)
) |> 
  collect_predictions()

cal_plot_regression(
  ridge_pred_check,
  truth = fuel_load_Mg_ha,
  estimate = .pred
)
```

this actually looks really bad... we are essentially predicting the same value no matter what?

```{r}
ridge_fit <- ridge_wflow_final |>
  fit(data = fd_train)
ridge_fit
```

```{r}
# look at the ridge path
ridge_fit |>
  extract_fit_engine() |>
  plot(xvar = "lambda", label = TRUE)
```

```{r}
# look at coefficient estimates
ridge_coef <- ridge_fit |>
  broom::tidy()
ridge_coef
```

```{r}
# look at predictions
predictions_ridge <- ridge_fit |>
  broom::augment(new_data = fd_test)
predictions_ridge |>
  dplyr::select(
    fuel_load_Mg_ha,
    .pred,
    TreesN,
    MeanTH,
    Basalarea,
    everything()
  )
rmse(predictions_ridge, truth = fuel_load_Mg_ha, estimate = .pred)
```


## LASSO model
```{r}
# lasso model
lasso_model <- linear_reg(mode = "regression", engine = "glmnet",
                         penalty = tune(),
                         mixture = 1)

lasso_wflow <- workflow() |>
  add_model(lasso_model) |>
  add_recipe(ridge_recipe) # same recipe as before
```

```{r}
# tune the model
lasso_tune1 <- tune_grid(lasso_model,
                         ridge_recipe,
                         resamples = fd_cv,
                         grid = grid_regular(penalty(range = c(-5, 5)), levels= 50))

```

```{r}
# plotting lasso penalty values
lasso_tune1 |>
  collect_metrics() |>
  filter(.metric == "rmse") |>
  ggplot(mapping = aes(x = penalty, y = mean)) + geom_point() + geom_line() +
  scale_x_log10()
```


```{r}
lasso_best <- lasso_tune1 |>
  select_best(metric = "rmse")

lasso_best

lasso_wflow_final <- lasso_wflow |>
  finalize_workflow(parameters = lasso_best) 
```

```{r}
# callibration check
lasso_pred_check <- lasso_wflow_final |>
  fit_resamples(
    resamples = fd_cv,
    control = control_resamples(save_pred = TRUE)
) |> 
  collect_predictions()

cal_plot_regression(
  lasso_pred_check,
  truth = fuel_load_Mg_ha,
  estimate = .pred
)
```
this at least looks a little better than before, there is a least a little bit of a slope...

```{r}
lasso_fit <- lasso_wflow_final |>
  fit(data = fd_train)
lasso_fit
```

```{r}
# look at lasso path
lasso_fit |>
  extract_fit_engine() |>
  plot(xvar = "lambda", label = TRUE)
```

```{r}
# lasso coefficients
lasso_coef <- lasso_fit |>
  broom::tidy()
lasso_coef 
```
```{r}
# look at predictions
predictions_lasso <- lasso_fit |>
  broom::augment(new_data = fd_test)
predictions_lasso |>
  dplyr::select(
    fuel_load_Mg_ha,
    .pred,
    TreesN,
    MeanTH,
    Basalarea,
    everything()
  )
rmse(predictions_lasso, truth = fuel_load_Mg_ha, estimate = .pred)
```

